---
title: "prediction analysis"
output: html_document
---
Runtime: 7‘56” in AWS
# Preparations

## Package Loading

```{r}
library(tidyverse)
library(haven)
```

## Data Loading

```{r}
df1 <- read_csv("https://raw.githubusercontent.com/celesteisyy/Economic-Analysis-In-R/refs/heads/main/FinalProject550/historic_property_data.csv")
head(df1)
```

## Check Data
First check overall data

```{r}
str(df1)
#make sure to collapse the output after viewing the info for bettering viewing
```

### Data Type Conversion
Notice that the type of `ind_large_home`, `ind_garage` and `ind_arms_length`(which is not mentioned in the codebook, quite confusing) is *logi*, we need to convert them into categorical type for models to interpret:

```{r}
# Convert logical data to factor
df1$ind_large_home <- as.factor(df1$ind_large_home)
df1$ind_garage <- as.factor(df1$ind_garage)
df1$ind_arms_length <- as.factor(df1$ind_arms_length)
```


### NA Omision and Substitute

Here, I manually dropped `meta_cdu`, `char_apts`, `char_attic_fnsh`, `char_renovation`, `char_porch` due to **the large amount of NAs**, which made these variables unavailable for analysis. I also dropped `char_repair_cnd`, `char_site`,`char_cnst_qlty` **based on the codebook guidance**.

**They might influence the final outcome, but the number of available rows was too small to analyze.**

```{r}
df1_1 <- df1 %>%
  select(-meta_cdu, -char_apts, -char_attic_fnsh, -char_renovation, -char_porch)%>%
  select(-char_repair_cnd, -char_site, -char_cnst_qlty)

head(df1_1)
```

In the simplest case (not considering other, more effective but fancier ML methods such as KNN), to deal with NAs, we fill the them with the **Random Imputation** based on their distribution, since we don't want to change the original distribution.
```{r}
set.seed(550)  # Set a random seed to ensure reproducibility
# Create a copy of df1_1 to store the cleaned data
df1_cleaned <- df1_1

# Iterate through all columns in df1_cleaned
for (var in names(df1_cleaned)) {
  # Check if the column contains NA values
  if (any(is.na(df1_cleaned[[var]]))) {
    # Randomly sample from non-NA values to fill the missing values
    df1_cleaned[[var]][is.na(df1_cleaned[[var]])] <- sample(
      df1_cleaned[[var]][!is.na(df1_cleaned[[var]])],  # Non-NA values to sample from
      sum(is.na(df1_cleaned[[var]])),                 # Number of NA values to fill
      replace = TRUE                                  # Allow repeated sampling
    )
  }
}
```

## Check Again if Data is Clean

Finally, let's check the cleaned data:
```{r}
df1_cleaned_na <- data.frame(colSums(is.na(df1_cleaned)))
sum(df1_cleaned_na)
```

## Scaling Data

Since large-scale differences in our data may be caused by extreme outliers, scaling might stabilize model behavior.
```{r}
df1_standardized <- df1_cleaned

# Create a list to store scaling parameters
scaling_params <- list()

# Standardize numeric columns and save scaling parameters
for (var in names(df1_standardized)) {
  if (is.numeric(df1_standardized[[var]])) {
    # Save mean and standard deviation
    scaling_params[[var]] <- list(
      mean = mean(df1_standardized[[var]], na.rm = TRUE),
      sd = sd(df1_standardized[[var]], na.rm = TRUE)
    )
    # Standardize the column
    df1_standardized[[var]] <- (df1_standardized[[var]] - scaling_params[[var]]$mean) / scaling_params[[var]]$sd
  }
}
```

Phew, good enough to proceed!

# Modelling 1 - Feature Selection

Here we first choose *Random Forest* for more precise feature selection

## Packge Loading
```{r}
#install.packages("randomForest")
library(randomForest)
```

## Feature Selecting

```{r}
set.seed(550)
rf_feature <- randomForest(
  sale_price ~ ., 
  data = df1_standardized, 
  keep.inbag = TRUE, # keep selected feature during tree-forming process
  ntree = 15, 
  importance = TRUE
)

print(rf_feature)
```

We select features based on their importance. Keep only variable whose %IncMSE is greater than 2
```{r}
# Extract importance scores
scores <- importance(rf_feature)

# Sort variables by `%IncMSE` in descending order
sorted_scores <- scores[order(scores[, "%IncMSE"], decreasing = TRUE), ]

# Set a threshold for `%IncMSE` (e.g., importance > 1.0)
threshold <- 2
selected_vars <- rownames(sorted_scores)[sorted_scores[, "%IncMSE"] > threshold]

# Print the selected variables
print("Variables with Importance Above Threshold (%IncMSE):")
print(selected_vars)
```

Before the second model is trained, we need to store our selected features:
```{r}
df1_selected <- df1_standardized[, c(selected_vars, "sale_price")]
```


# Modelling 2 - Price Prediction

In this part, we need to use a `predict_property_data` for price predicting

## Data Loading and Cleaning
Same process as loading df1:
```{r}
df2 <- read_csv("https://raw.githubusercontent.com/celesteisyy/Economic-Analysis-In-R/refs/heads/main/FinalProject550/predict_property_data.csv")
head(df2)
```

Clean out NAs in df2:
```{r}
# Convert logical data to factor
df2$ind_large_home <- as.factor(df2$ind_large_home)
df2$ind_garage <- as.factor(df2$ind_garage)
df2$ind_arms_length <- as.factor(df2$ind_arms_length)
```

```{r}
df2_1 <- df2 %>%
  select(-meta_cdu, -char_apts, -char_attic_fnsh, -char_renovation, -char_porch)%>%
  select(-char_repair_cnd, -char_site, -char_cnst_qlty)

head(df2_1)
```

Also, we use **Random Imputation** method to fill NAs.
```{r}
set.seed(550)

# Create a copy of df2_1 to store the cleaned data
df2_cleaned <- df2_1

# Iterate through all columns in df2_cleaned
for (var in names(df2_cleaned)) {
  # Check if the column contains NA values
  if (any(is.na(df2_cleaned[[var]]))) {
    # Randomly sample from non-NA values to fill the missing values
    df2_cleaned[[var]][is.na(df2_cleaned[[var]])] <- sample(
      df2_cleaned[[var]][!is.na(df2_cleaned[[var]])],  # Non-NA values to sample from
      sum(is.na(df2_cleaned[[var]])),                 # Number of NA values to fill
      replace = TRUE                                  # Allow repeated sampling
    )
  }
}
```


```{r}
df2_cleaned_na <- data.frame(colSums(is.na(df2_cleaned)))
sum(df2_cleaned_na)
```
```{r}
df2_standardized <- df2_cleaned

for (var in names(df2_standardized)) {
  if (is.numeric(df2_standardized[[var]]) && !is.null(scaling_params[[var]])) {
    # Standardize using the saved mean and standard deviation
    df2_standardized[[var]] <- (df2_standardized[[var]] - scaling_params[[var]]$mean) / scaling_params[[var]]$sd
  }
}
```

## Extract Selected Features

```{r}
# Check if df2 contains all the required variables
missing_vars <- setdiff(selected_vars, names(df1_selected))
if (length(missing_vars) > 0) {
  stop(paste("The following variables are missing in df2:", paste(missing_vars, collapse = ", ")))
}

# Subset df2 based on the selected variables
df2_selected <- df2_standardized[, selected_vars]
```

## Model Constructing

First we need to train the *Second Random Forest* Model for Prediction

### Model Training
```{r}
set.seed(550)
rf_predict <- randomForest(
  sale_price ~ ., 
  data = df1_selected, 
  ntree = 35,
  importance = TRUE
)
```

### Final MSE
```{r}
print(rf_predict)
print(mean(rf_predict$mse))
```

### Predicting

```{r}
# Predict on standardized df2
pred_std <- predict(rf_predict, newdata = df2_selected)

# Reverse standardization for sale_price predictions
pred<- (pred_std * scaling_params[["sale_price"]]$sd) +
  scaling_params[["sale_price"]]$mean

# Add predictions to df2
df2$predicted_sale_price <- pred

# View the final results
head(df2$predicted_sale_price)
```

# Backtesting Data
Same process as we've done in `predict_property_data`, here we use the entire df1 to test our outcome
```{r}
# Predict on standardized df2
pred_df1 <- predict(rf_predict, newdata = df1_standardized)

# Reverse standardization for sale_price predictions
pred1<- (pred_df1 * scaling_params[["sale_price"]]$sd) +
  scaling_params[["sale_price"]]$mean

# Add predictions to df2
df1$predicted_sale_price <- pred1

# View the final results
head(df1)
```
# Summary

## Summary Statistics

```{r}
summary(df2$predicted_sale_price)
```
```{r}
# Create the assessment dataframe
assessment_df <- data.frame(
  pid = df2$pid,                      # Property identification numbers
  assessed_value = df2$predicted_sale_price  # Predicted property values
)

# Ensure all values are non-missing and non-negative
assessment_df <- assessment_df[!is.na(assessment_df$assessed_value) & assessment_df$assessed_value >= 0, ]

# Preview the assessment dataframe
head(assessment_df)
```


# Data Output

Save the file in current working directory
```{r}
# Export to a CSV file
write.csv(assessment_df, "assessed_value.csv", row.names = FALSE)
```
