---
title: "lab 9"
output: html_document
date: "2023-11-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Inequality in Spain

```{r libraries}
library(tidyverse)
library(readr)
library(forcats)
```

```{r National Income}
df <-  read_delim("data/income.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE, skip = 1)%>%
  rename(percentile = Percentile,
         date = Year, 
         value = `sptinc_z_ES\nPre-tax national income \nTop 10% | share\nSpain`)
```

## Plots

```{r Pre-tax, echo=FALSE}
p1<-ggplot( df , aes(x = date, y = value, shape = percentile))+
  geom_line(lwd = 1)+
  geom_point(size=3)+
  xlab("") +
  ylab("Share of national income (%)") +
  labs(title = "Income Inequality in Spain, 2004-2020",
       caption = "Source: World Inequality Database")+
  theme(plot.title = element_text(size=15),
        legend.title = element_blank(),
        legend.text=element_text(size=10),
        legend.position="top",
        plot.caption = element_text(size=9, hjust = 0),
        axis.title = element_text(size = 13),
        strip.background = element_blank(),
        axis.text.x=element_text(size = 13, hjust = 0.7, angle = 45),
        axis.text.y=element_text(size = 10),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill='transparent'),
        legend.key=element_blank())+
  scale_x_continuous(n.breaks = 6)

```

##Gini Index--
I used Gini index from 1980 to 2020 for Spain as dataset and drew the plot of its development:
```{r Gini in Spain}
evo<-read_csv("data/gini.csv", col_types = cols(`1960` = col_double(), 
                                                `1961` = col_double(), `1962` = col_double(), 
                                                `1963` = col_double(), `1964` = col_double(), 
                                                `1965` = col_double(), `1966` = col_double(), 
                                                `1967` = col_double(), `1968` = col_double(), 
                                                `1969` = col_double(), `1970` = col_double(), 
                                                `1972` = col_double(), `1974` = col_double(), 
                                                `1976` = col_double(), `1978` = col_double(), 
                                                ...68 = col_double()), skip = 4)%>%
  select(-'Country Code',-'Indicator Name',-'Indicator Code')%>%
  rename(country='Country Name')%>%
  filter(country=='Spain')%>%
  gather(year,gini,2:64)%>%
  mutate(year=as.numeric(year))%>%
  filter(!is.na(gini))%>%
  select('country','year','gini')%>%
  filter(year>=2004)
```

Here comes the plot:
```{r Plotting Gini in Spain}
p2 <- ggplot(evo, aes(x = year, y = 100*(gini/33.3)))+
  geom_line(lwd = 1)+
  theme(plot.title = element_text(size=15),
        plot.subtitle = element_text(size=9),
        legend.title = element_blank(),
        legend.position="bottom",
        plot.caption = element_text(size=9, hjust = 0),
        axis.title = element_text(size = 13),
        strip.background = element_blank(),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        legend.key = element_rect(fill = NA),
        text=element_text(family="Palatino"),
        axis.text.x=element_text(size=13,angle=45, hjust=1),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))+
  xlab("") +
  ylab("Gini Index (2004=100)") +
  labs(title = "",
       caption = "Source: OECD, Short-term Unemployment Statistics and World Bank, World Development Indicators.")+
  scale_x_continuous(n.breaks = 6)
```

##Unemployment Rate:
What about combining pre-tax national income with unemployment rate?
```{r unemployment}
data<-read_csv('data/labor.csv')%>%
  select(TIME, Measure, Value) %>%
  rename(date  = TIME) %>%
  rename(adj   = Measure) %>%
  rename(unemp = Value) %>%
  mutate(adj  = ifelse(adj == 'Level, rate or quantity series', 'No seasonal adjustment', 'Seasonal adjustment')) %>%
  mutate(date = paste(date, "01", sep="-")) %>%
  mutate(date = as.Date(date, format = "%Y-%m-%d"))
```
Then plot it:
```{r plotting unemployment}
p3<-ggplot(data, aes(x = date, y = unemp, color = adj))+
  geom_line(stat='identity',lwd=1)+
  theme(plot.title = element_text(size=15),
        legend.title = element_blank(),
        legend.position="bottom",
        plot.caption = element_text(size=10, hjust = 0),
        plot.subtitle = element_text(size=10, hjust = 0),
        axis.title = element_text(size = 11),
        strip.background = element_blank(),
        axis.text.x=element_text(angle=45, hjust=1),
        strip.placement = "outside",
        panel.grid.minor = element_line('grey', size = 0.1, linetype = 'dashed'),
        panel.grid.major = element_line('grey', size = 0.1, linetype = 'dashed'),
        legend.key = element_rect(fill = NA),
        text=element_text(family="Palatino"),
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))+
  scale_x_date(date_breaks = "12 month", date_labels = "%m-%Y")+
  xlab("") +
  ylab("Unemployment rate (%)") +
  labs(title = "Unemployment rate in Spain",
       subtitle='From Jan. 2004 to Dec. 2020',
       caption = "")
```


Then combined the pre-tax index with pre-tax national income per share in a single graph:
```{r combined graph}
library(ggpubr)
ggarrange(p1, p2, ncol=1, nrow = 2)
```

```{r}
cowplot::plot_grid(p3, p2, nrow = 2, rel_heights = c(0.58, 0.42), align = "v")
```


The above figure plots the pre-tax national income per share as well as the development of Gini index in Spain from 1980 to 2022, constructed by the dataset from World Inequality Database.(WID,2023) The overall trend of the income inequality is stable throughout the year.

Looking closer the dataset, from 1986 to 1991, the proportion of pre-tax national income of the top 10% in the country reached to the level of roughly 38%. In year 1989, the pre-tax national income in Spain witnessed the highest value of 38.63% of the top 10% income group, mostly attribute by the income increase in the top 1% within the county, meaning the overall income in the country increased during the time.

However, from 1990 to 2022, the trend started to overturned. The percentage in the top 50% group declined along with the rise of that in the bottom 50% group, the latter eventually stabled at the level of 20%. This suggests that the overall inequality decreased during the time.




